---
title: "Parent_Capstone_Project"
author: Alexis Parent 
date: 11/1/2024
format: html
editor: visual
---
In this qmd file I will be exploring and cleaning data in the csv files "fig_S9.csv" and "fig_17.csv". 

```{r}
# clear R's brain and set up R 
rm(list=ls())
library(tidyverse)
library(multcomp)
library(dplyr)
library(ggfortify)
library(here)
```

```{r}
# read in csv file for both datasets
organic_matter_soil <- read.csv(here("Parent-Capstone-Project", "Data", "fig_S9.csv"))

total_N <- read.csv(here("Parent-Capstone-Project", "Data", "fig_S17.csv"))
```

Next we are point to take a look at the data: 

```{r}
# take a look at the data 
glimpse(total_N)
levels(total_N)

glimpse(organic_matter_soil)
levels(organic_matter_soil)
```

From the total_N data, we can see there are 36 rows and 23 columns. Most of which I will not be utilizing except, SOC and total_N. Both of which are numerical. 

From the organic_matter_soil data, we can see there are 24 rows and 5 columns. I will be utilizing the columns rotation, POM and MAOM.

```{r}
# isolate columns needed for data analysis for total_N dataset
SOC_total_N <- total_N |> dplyr::select(SOC,total_N)

print(SOC_total_N)
view(SOC_total_N)

# isolate columns for organic_matter dataset
organic_matter <- organic_matter_soil |> dplyr::select(rotation, POM, MAOM)
view(organic_matter)
```

Now that the data is isolated into the columns I will be looking at, I can now check the levels of the variables to make sure there are no mistakes. 

```{r}
# check levels for SOC_total_N 
levels(SOC)
# check levels for organic_matter
levels(organic_matter)
```
After checking the levels we can see there are 

The code below is going to identify outliers for both datasets

```{r}
# identify outliers for SOC_total_N 
ggplot(SOC_total_N, aes(SOC)) + 
  geom_histogram()

# identify outliers using for organic matter 
ggplot(organic_matter, aes(MAOM)) + 
    geom_histogram() 

ggplot(organic_matter, aes(POM)) + 
    geom_histogram() 
 
```

Interpretation:

In SOC_total_N there seems to be one outlier on the right hand side with a high SOC and a high total_N concentration. However, I am going to leave this outlier in the data because it can be indicative that as SOC increases total N will also continue to increase beyond where the other points are clustered in the graph. The outlier could also be indicitive of a unique observation about how high total nitrogen is compared to the amount of SOC.

In the organic_matter data set, there don't seem to be outliers in either the MAOM and POM. For POM, the bars are more equally distributed between 0 and 0.25 with a few bars on either side. Furthermore, for MAOM the bar plot is more left skewed with only a few values around 1.0 and most values around 0 or below. 

Furthermore, there are negative values for both and although I believe they are accurate, I do not think they are appropriate for the data set and should get rid of them (left a comment for you in teams to discuss!!)

Next we are going to save the cleaned data sets: 

```{r}
# cleaned data 
# define file path
data_folder_path <- "Users/alexisparent/Desktop/BIOSTAT/Parent-Capstone-Project/Data"

# cleaned files
organic_matter_clean.csv <- "organic_matter"
SOC_total_N_clean.csv <- "SOC_total_N"

#combine file path and names of csv files
file_path <- file.path(data_folder_path, "organic_matter_clean.csv", "SOC_total_N_clean.csv")

# save cleaned dataset as csv files
write.csv(organic_matter, file = "organic_matter_clean.csv", row.names = FALSE)
write.csv(SOC_total_N, file = "SOC_total_N_clean.csv", row.names = FALSE)

```

Now that the data is cleaned and saved into the data folder, we can now perform an exploratory data analysis. The code below begins the exploratory analysis.

```{r}
# exploratory analysis 
# mean and sd for organic_matter_soil for POM
organic_matter_soil |> group_by(rotation) |> summarise(
  mean_POM = mean(POM, na.rm = T),
  sd_POM= sd(POM, na.rm = T))

 # mean and sd for MAOM
organic_matter_soil |> group_by(rotation) |> summarise(
  mean_MAOM = mean(MAOM, na.rm = T),
  sd_MAOM= sd(MAOM, na.rm = T))
```

Organic matter soil: From the mean of POM, we can see that as years (rotation) go on the mean POM increases. Furthermore, as the rotation increases in years for MAOM, the mean MAOM also increases greatly from year 2 to 3 and then decreases from year 3 to 4. As a result, it looks like there could be a significant relationship between number of years rotated and the amount of POM/MAOM in the soil.

Exploratory Analysis: 

```{r}
# mean and sd for total_N
SOC_total_N |> group_by(SOC) |> summarise(
  mean_total_N = mean(total_N, na.rm = T),
  sd_total_N= sd(total_N, na.rm = T))
```

SOC_total_N: As SOC continues to increase, the mean total nitrogen concentration also continues to increase. Until the end where it hits 2 and then decreases just slightly as SOC goes up.

As a result from the graphs above and looking at the relationships within both sets of data. I believe a regression model would be the most appropriate for SOC_total_N. The response variable is total_N in the soil, which is ratio data, with a continuous predictor variable (SOC). There is only one predictor variable (SOC), as a result, this leads me to a linear regression model. 

For the organic_matter_soil data set we are looking at SOC and MAOM as a response variable for rotation (years). SOC and MAOM are both continuous response variables, and rotation (years) is a categorical varibale. Which I believe would be a regression test. 



