---
title: "total_N_SOC"
author: Alexis Parent 
date: 11/11/2024
format: html
editor: visual
---

I am going to be looking at the relationship between total Nitrogen in the soil and SOC in the soil using the data set 'SOC_total_N_clean.csv'. In this preliminary data analysis I will be conducting a linear regression. The abbreviation SOC represents soil organic carbon and total N represents the total Nitrogen in the soil.

Null Hypothesis: I believe there is no relationship between soil organic carbon and total nitrogen in the soil. 

Alternative Hypothesis: I believe there will be a positive linear relationship between soil organic carbon and nitrogen in the soil. 

```{r}
# set up R and clear R's brain
rm(list=ls())
library(tidyverse)
library(multcomp)
library(dplyr)
library(ggfortify)
library(here)
```

```{r}
# read in cleaned file 
total_N_SOC <-  read_csv(here("Data", "SOC_total_N_clean.csv"))
```

```{r}
# look at data
glimpse(total_N_SOC)
view(total_N_SOC)
```

From the total_N data, we can see there are 36 rows and 2 columns (SOC, total_N). I used glimpse so I was able to quickly view what the data looked like. I then used view so I was able to open up another tab of the total data to see everything as a whole.

```{r}
# find the mean and sd of total N
total_N_SOC|> summarise(
  mean_total_N = mean(total_N, na.rm = T),
  sd_total_N = sd(total_N, na.rm = T))

# find the mean and sd of SOC 
total_N_SOC |> summarise(
  mean_SOC = mean(SOC, na.rm = T), 
  sd_SOC = sd(SOC, na.rm = T))
```

The mean for total_N concentration is 1.749. The standard deviation for total_N concentration is 0.2394. Which means the values of N concentration deviate around 0.2394 from 1.749.

Furthermore, the mean for SOC is 20.82 and the standard deviation is 4.138133.


```{r}
# plot the relationships 
ggplot(total_N_SOC, aes(x = total_N, y = SOC)) +
  geom_point(color = "darkgreen", size = 2)
```

From the scatter plot we can see there is a positive linear relationship between SOC in the soil and total nitrogen in the soil. As total N increases, SOC also increases.

Guesstimate: I believe there will be a linear slope between total N and SOC. 
and the y intercept will be 

Next we are going to make the model and check assumptions:

```{r}
# make a model 
N_SOC_mod <- lm(total_N ~ SOC, data = total_N_SOC)
```

```{r}
# check model assumptions
autoplot(N_SOC_mod, smooth.colour = NA)
```

From the assumptions we can see that the observations follow the line of normal distribution. Which allows us to go ahead and conduct further tests.

Next we will conduct a regression model to see if there is a significant difference between SOC and total N concentration in the soil.

```{r}
# conduct test for regression model
summary(N_SOC_mod)
```

Interpretation:

A linear regression was conducted to see if there was an effect between soil organic carbon (SOC) and total Nitrogen (total_N) concentration in the soil. From the results we can see there is an effect between total nitrogen and SOC concentration. The adjusted R-square for the number of predictors in the model indicates approximately 85.8% of the variance in the data (p \< 0.05, f = 212.6, df = 34, R^2 = 0.8621). 

Next we will make a clean figure.

```{r}
# making a clean model
ggplot(total_N_SOC, aes(x = total_N, y = SOC)) +
  geom_point(color = "darkgreen", size = 2) + 
  labs( 
    title = "Relationship Between SOC and Total N in Soil",
    x = "Total Nitrogen in Soil",
    y = "SOC")
```

More Info: 

Intercept:  When SOC is 0 the estimated value of total N is 0.630839. 
The slope for SOC is 0.053719. Therefore, for each unit increase in SOC, total N incrwases by 0.053719. 
